Testing for measurement invariance with respect to an ordinal variable

### 2012-08-09 ordinal/ordinal.Rnw
Added Simulation 2 method, results to ordinal paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@157 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<preliminaries>>=
## packages
library("lavaan")
library("strucchange")
library("mvtnorm")
library("lattice")

@@

## convenience function for plotting boundaries without color
get_boundary <- function(obj, fun, trim = 0) {
  bound <- fun$boundary(0:obj$nobs/obj$nobs)
  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
  bound <- zoo(bound, time(obj$process))
  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
  return(bound)
}
@

@@

if(file.exists("sim1.rda")) {
  load("sim1.rda")
} else {
  sim1 <- simulation()
  save(sim1, file="sim1.rda")
}
@

@@

Simulation results are presented in Figure XX.  Rows of the figure
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_O$ statistic from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see Merkle and Zeileis, 2012, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than .05 at $d=0$).  This is because the likelihood
ratio test requires 
estimation of a multiple groups model, which is very unstable with
large numbers of groups and small sample sizes.  The proposed
statistics, which require estimation of only the single-group model,
exhibit a clear advantage in these conditions.

@@

In this simulation, we demonstrate that the proposed statistic is
relatively insensitive to minor parameter violations that are
unrelated to the ordering of the auxiliary variable.  As noted
earlier, this feature is especially applicable to situations where
one's sample size is very large.  Analysts often resort to informal
fit measures in this case, because the traditional LRT is nearly
guaranteed to result in significance.  This simulation is intended to
show that the proposed ordinal test remains viable for very large $n$.

<<sim2>>=
if(file.exists("sim2.rda")) {
  load("sim2.rda")
} else {
  sim2 <- simulation(diff = seq(0, 0.5, by = 0.1), nobs = c(1200, 4800, 9600),
                     anomaly=TRUE)
  save(sim2, file="sim2.rda")
}
@

@@

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=1:4)
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ordL2BB(gefp1))

@@

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=3:6)
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ordL2BB(gefp2))

@@

%% application to other psychometric models

%% should we care about non-ordinal violations when using an ordinal variable?

>>>>

<<preliminaries>>=
## packages
library("lavaan")
library("strucchange")
library("mvtnorm")
library("lattice")
library("foreign")

@@

## convenience function for plotting boundaries without color
get_boundary <- function(obj, fun, trim = 0) {
  bound <- fun$boundary(0:obj$nobs/obj$nobs)
  bound <- fun$computeCritval(0.05, ncol(obj$process)) * bound
  bound <- zoo(bound, time(obj$process))
  if(trim > 0) bound <- head(tail(bound, - floor(trim * obj$nobs)), - floor(trim * obj$nobs))
  return(bound)
}

## convenience function for applying tests to real data
info_full <- function(x, ...) solve(vcov(x) * nobs(x))
@

@@

if(file.exists("sim1.rda")) {
  load("sim1.rda")
} else {
  sim1 <- simulation()
  sim1$nlevels <- factor(sim1$nlevels)
  levels(sim1$nlevels) <- paste("m=",levels(sim1$nlevels),sep="")
  sim1$nobs <- factor(sim1$nobs)
  levels(sim1$nobs) <- paste("n=",levels(sim1$nobs),sep="")
  levels(sim1$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
  save(sim1, file="sim1.rda")
}
@

@@

Simulation results are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_O$ statistic from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see Merkle and Zeileis, 2012, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than .05 at $d=0$).  This is because the likelihood
ratio test requires 
estimation of a multiple groups model, which is very unstable with
large numbers of groups and small sample sizes.  The proposed
statistics, which require estimation of only the single-group model,
exhibit a clear advantage in these conditions.

@@

In this simulation, we demonstrate that the proposed statistic is
relatively insensitive to minor parameter violations that are
unrelated to the ordering of the auxiliary variable.  As noted
earlier, this feature is especially applicable to situations where
one's sample size is very large.  Analysts often resort to informal
fit measures in this case, because the traditional LRT is nearly
guaranteed to result in significance.  This simulation is intended to
show that the proposed ordinal test remains viable for large $n$.

\subsection{Method}
Data were generated from the same factor analysis model that was used
in Simulation 1, with measurement invariance violations in the unique
variance parameters.  To implement a minor measurement invariance
violation, the unique variances were equal across all levels of the
ordinal variable except one (level $1+m/2$).  At this particular
level, the unique variances were greater by a factor of $d$~times the
parameters' asymptotic standard errors (scaled by $\sqrt{n}$), with
$d=0, 0.5, 1.0, \dots, 3.0$.  The number of levels of the ordinal
variable were the same as those in Simulation 1 ($m=4,8,12$), and
sample sizes were set at $n=1200,4800,9600$.  All other simulation
features matches those from Simulation 1.

<<sim2>>=
if(file.exists("sim2.rda")) {
  load("sim2.rda")
} else {
  sim2 <- simulation(diff = seq(0, 3, by = 0.5), nobs = c(1200, 4800, 9600),
                     anomaly=TRUE)
  sim2$nlevels <- factor(sim2$nlevels)
  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
  sim2$nobs <- factor(sim2$nobs)
  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
  levels(sim2$test) <- c("LM_uo", "LRT", "LM_o", "WDM_o")
  save(sim2, file="sim2.rda")
}
@

\subsection{Results}
Simulation results are presented in Figure~\ref{fig:sim2res}.  It is
observed that results are very consistent across the sample sizes
tested, implying that "practical infinity" is reached for this model
by $n=1200$. \readme{Also implying that we don't need all three sample
  sizes or could try values smaller than 1200.}  We also observe
a negative 
relationship between power and $m$; this is because the measurement
invariance violation occurred at only one level of the auxiliary
variable.  As $m$ increases (and $n$ is held
constant), the number of individuals violating measurement invariance
therefore decreases.  As a result, power to detect the violation decreases with
increasing $m$.

The more interesting result of this simulation lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the finding
of Bentler and Bonett \citeyear{BenBon80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\text{WDM}_O$ statistic being the lowest and the $\text{LM}_O$
statistic being higher at larger values of $d$.  
This demonstrates that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation occurring at a single level of the auxiliary
variable.

\begin{figure}
\caption{Simulated power curves for the ordered and unordered
      $\max \mathit{LM}$ tests, the ordered double-max test, and the
      likelihood ratio test across three sample sizes $n$, three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--3 standard errors (scaled by $\sqrt{n}$)
      occurring at a single level (the $(1 +
      m/2)^{\text{th}}$ level) of the ordinal auxiliary variable.}
\label{fig:sim2res}
\setkeys{Gin}{width=\textwidth}
<<sim2res, fig=TRUE, height=7, width=7>>=
mykey <- simpleKey(levels(sim2$test), points = FALSE, lines = TRUE)
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@
\end{figure}

Taken together, the results from Simulation 1 and Simulation
2 provide evidence that the $\text{LM}_O$ statistic should be
preferred to the $\text{WDM}_O$ statistic for testing measurement
invariance.  The $\text{LM}_O$ statistic has higher power to detect
ordinal violations, and its power to detect non-ordinal violations was
similar to that of $\text{WDM}_O$ when the violation magnitude was
small (e.g., for $d \leq 1.5$).  The $\text{LM}_O$ statistic is
advantageous because it can make use of invariance violations that
simultaneously 
occur in multiple parameters, whereas the $\text{WDM}_O$ focuses only
on the parameter with the largest invariance violation.  
Thus, the statistics are likely to exhibit
similar performance if only a single model parameter violated
measurement invariance.

In the next section, we compare the proposed statistics to the
likelihood ratio test with real data.

@@

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=1:4)
if(file.exists("ol2bb1.rda")){
  load("ol2bb1.rda")
} else {
  ol2bb1.fun <- ordL2BB(gefp1)
  save(ol2bb1.fun, file="ol2bb1.rda")
}
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)

@@

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=3:6)
if(file.exists("ol2bb2.rda")){
  load("ol2bb2.rda")
} else {
  ol2bb2.fun <- ordL2BB(gefp2)
  save(ol2bb2.fun, file="ol2bb2.rda")
}
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)

@@

\begin{enumerate}
  \item The only proposed statistics that treat ordinal auxiliary
    variables as truly ordinal?  LRT treats them as categorical,
    inclusion in the model typically treats tham as continuous.
  \item The first point is important because, if we test wrt an
    ordinal auxiliary variable, we are generally looking for ordinal
    violations.  We don't care if a middle level differs from
    all others because we would not usually have an intuitive
    explanation for it.
  \item The proposed statistics should generally transfer to IRT and
    other psychometric models.
\end{enumerate}
